---
title: "Introduction to ltmle()"
author: "Joshua Schwab"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(ltmle)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Minimal Example
Consider the observed data with time ordering W -> A -> Y.

W ~ N(0, 1)

A ~ binomial with P(A = 1) = expit(W)

Y ~ binomial with P(Y = 1) = expit(W + A)

where $expit(z) = \frac{1}{1 + e^{-z}}$

We want to know $E[Y_1]$ the expected value of Y, intervening to set A to 1.
```{r}
set.seed(1234)
rexpit <- function(x) rbinom(n=length(x), size=1, prob=plogis(x))
n <- 10000
W <- rnorm(n)
A <- rexpit(W)
Y <- rexpit(W + A)
data <- data.frame(W, A, Y)
```

We could try to use the simple mean of Y or the mean of Y when A = 1, but these will be biased.
```{r}
mean(Y)
mean(Y[A == 1])
```

Now we use ltmle().
```{r}
result <- ltmle(data, Anodes = "A", Ynodes = "Y", abar = 1)
result
```
Because we're using simulated data, we can calculate the true value of $E[Y_1]$
```{r}
n.large <- 1e6
W <- rnorm(n.large)
A <- 1
Y <- rexpit(W + A)
mean(Y)
```

## Single time point 
Time ordering of data is W1 -> W2 -> W3 -> A -> Y
```{r}
n <- 1000
W1 <- rnorm(n)
W2 <- rbinom(n, size=1, prob=0.3)   
W3 <- rnorm(n)
A <- rexpit(-1 + 2 * W1 + W3)
Y <- rexpit(-0.5 + 2 * W1^2 + 0.5 * W2 - 0.5 * A + 0.2 * W3 * A - 1.1 * W3)
data <- data.frame(W1, W2, W3, A, Y)
```
True value of $E[Y_1]$ is approximately 0.5939.

SuperLearner semiparametric estimation using all parents as regressors 
``` {r}
result <- ltmle(data, Anodes="A", Lnodes=NULL, Ynodes="Y", abar=1, SL.library="default")
```

TMLE estimate:
```{r}
summary(result)
```
IPTW estimate:
```{r}
summary(result, estimator="iptw")
```

SuperLearner semiparametric estimation using correctly specified regressors
```{r}
result <- ltmle(data, Anodes="A", Lnodes=NULL, Ynodes="Y", Qform=c(Y="Q.kplus1 ~ I(W1^2) + W2 + W3*A"), gform="A ~ W1 + W3", abar=1, SL.library="default")
summary(result)
```

glm using correctly specified Qform and gform
```{r}
result <- ltmle(data, Anodes="A", Lnodes=NULL, Ynodes="Y", 
 Qform=c(Y="Q.kplus1 ~ I(W1^2) + W2 + W3*A"), gform="A ~ W1 + W3", 
 abar=1, SL.library=NULL)
summary(result)
```


Get summary measures (additive treatment effect, odds ratio, relative risk) for abar=1 vs abar=0
```{r}
result <- ltmle(data, Anodes="A", Lnodes=NULL, Ynodes="Y", 
                      abar=list(1, 0), SL.library="default")
summary(result)
```

## Censoring
W -> C -> Y
```{r}
n <- 100000
W <- rnorm(n)
C <- BinaryToCensoring(is.censored = rexpit(W))
summary(C)
Y <- rep(NA, n)
Y[C == "uncensored"] <- rexpit(W[C == "uncensored"])
data <- data.frame(W, C, Y)
head(data, 20)
result <- ltmle(data, Anodes = NULL, Cnodes = "C", Ynodes = "Y", abar = NULL)
summary(result)
```
The naive estimate is biased:
```{r}
mean(data$Y)
mean(data$Y, na.rm = T)
```

## Longitudinal data
W -> A1 -> L -> A2 -> Y

```{r}
n <- 1000
W <- rnorm(n)
A1 <- rexpit(W)
L <- 0.3 * W + 0.2 * A1 + rnorm(n)
A2 <- rexpit(W + A1 + L)
Y <- rexpit(W - 0.6 * A1 + L - 0.8 * A2)
data <- data.frame(W, A1, L, A2, Y)
```


Treatment regime of interest: set all A1 to 1, set all A2 to 0:
```{r}
result.1.0 <- ltmle(data, Anodes=c("A1", "A2"), Lnodes="L", Ynodes="Y", abar=c(1, 0))
result.1.0
```

## Dynamic treatment
Treatment regime of interest is: 
 Always treat at time 1 (A1 = 1), treat at at time 2 (A2 = 1) if L > 0
```{r}
abar <- matrix(nrow=n, ncol=2)
abar[, 1] <- 1
abar[, 2] <- L > 0

result.abar <- ltmle(data, Anodes=c("A1", "A2"), Lnodes="L", Ynodes="Y", abar=abar)
result.abar
```

The regime can also be specified as a rule function
```{r}
rule <- function(row) c(1, row["L"] > 0)

result.rule <- ltmle(data, Anodes=c("A1", "A2"), Lnodes="L", Ynodes="Y", rule=rule)
result.rule
```

Specfifying the regime using abar and using rule give the same result:
```{r}
summary(result.abar)
summary(result.rule)
```





## Variance estimation
A simple point treatment problem W, A, Y. But there is a positivity problem - for small values of W, Prob(A = 1) is very small.

The true parameter value, $E[Y_1]$ is approximately 0.697.

The true TMLE standard deviation is approximately 0.064. 

The true IPTW standard deviation is approximately 0.058.

```{r}
n <- 1000
W <- rnorm(n)
A <- rexpit(8 * W)
Y <- rexpit(W + A)
r1 <- ltmle(data.frame(W, A, Y), Anodes="A", Ynodes="Y", abar = 1, estimate.time=FALSE)
r2 <- ltmle(data.frame(W, A, Y), Anodes="A", Ynodes="Y", abar = 1, estimate.time=FALSE, 
 variance.method="ic")
r3 <- ltmle(data.frame(W, A, Y), Anodes="A", Ynodes="Y", abar = 1, estimate.time=FALSE, 
 variance.method="iptw")
print(summary(r1))
print(summary(r2))
print(summary(r3))
print(summary(r1, estimator="iptw"))
print(summary(r2, estimator="iptw")) #the same - variance.method only affects TMLE
print(summary(r3, estimator="iptw")) #the same - variance.method only affects TMLE
```

We can see that the values of g are very small.
```{r}
summary(r1$cum.g)
summary(r1$cum.g.unbounded)
head(data.frame(data, g = r1$cum.g, g.unbounded = r1$cum.g.unbounded), 20)
```

## Hierachical data and the id variable
```{r}
num.households <- 500
people.in.household <- round(runif(num.households, min = 1, max = 10))
length(people.in.household)
n <- sum(people.in.household) 
n
W.household <- rnorm(num.households)
length(W.household)
W.household.expanded <- rep(W.household, times = people.in.household)
W.indiv <- rnorm(n)
length(W.indiv)
A <- rexpit(1.5 * W.household.expanded + 0.4 * W.indiv)
Y <- rexpit(-1 + 2.3 * W.household.expanded - 0.6 * W.indiv + 1.2 * A)
```
id can be an integer, factor, or character (or any type that can be coerced to factor)
```{r}
id <- 1:num.households 
id.expanded <- rep(id, times = people.in.household)
data <- data.frame(W.household.expanded, W.indiv, A, Y)
head(cbind(id.expanded, data), 20)
```
```{r}
result.without.id <- ltmle(data, Anodes = "A", Ynodes = "Y", abar = 0)
result.with.id <- ltmle(data, Anodes = "A", Ynodes = "Y", abar = 0, id = id.expanded)
```
Omitting the id variable makes the individuals seem more independent than they are, which gives artificially low variance estimates.
```{r}
summary(result.without.id)
summary(result.with.id)
```
The influence curve is a vector with length equal to the number of independent units.
```{r}
length(result.without.id$IC$tmle)
length(result.with.id$IC$tmle)
```

## Multiple time-dependent covariates and treatments at each time point, continuous Y values
age -> gender -> A1 -> L1a -> L1b -> Y1 -> A2 -> L2a -> L2b -> Y2
``` {r}
set.seed(2)
n <- 1000
age <- rbinom(n, 1, 0.5)
gender <- rbinom(n, 1, 0.5)
A1 <- rexpit(age + gender)
L1a <- 2*age - 3*gender + 2*A1 + rnorm(n)
L1b <- rexpit(age + 1.5*gender - A1)
Y1 <- plogis(age - gender + L1a + 0.7*L1b + A1 + rnorm(n))
A2 <- rexpit(age + gender + A1 - L1a - L1b)
L2a <- 2*age - 3*gender + 2*A1 + A2 + rnorm(n)
L2b <- rexpit(age + 1.5*gender - A1 - A2)
Y2 <- plogis(age - gender + L1a + L1b + A1 + 1.8*A2 + rnorm(n))
data <- data.frame(age, gender, A1, L1a, L1b, Y1, A2, L2a, L2b, Y2)
```

Also show some different ways of specifying the nodes:

```{r}
result5a <- ltmle(data, Anodes=c(3, 7), Lnodes=c("L1a", "L1b", "L2a", "L2b"), 
 Ynodes=grep("^Y", names(data)), abar=c(1, 0)) 
summary(result5a)
```

Usually you would specify a Qform for all of the Lnodes and Ynodes but in this case L1a, L1b, Y1 is a "block" of L/Y nodes not separated by Anodes or Cnodes (the same is true for L2a, L2b, Y2). Only one regression is required at the first L/Y node in a block. You can pass regression formulas for the other L/Y nodes, but they'll be ignored.
```{r}
result5b <- ltmle(data, Anodes=c(3, 7), Lnodes=c("L1a", "L1b", "L2a", "L2b"), 
 Ynodes=grep("^Y", names(data)), abar=c(1, 0), 
 Qform=c(L1a="Q.kplus1 ~ 1", L2a="Q.kplus1 ~ 1"))
summary(result5b)
```

Gives the same result but prints a message saying some regression formulas will be dropped:
```{r}
result5c <- ltmle(data, Anodes=c(3, 7), Lnodes=c("L1a", "L1b", "L2a", "L2b"), 
 Ynodes=grep("^Y", names(data)), abar=c(1, 0), 
 Qform=c(L1a="Q.kplus1 ~ 1", L1b="Q.klus1~A1", 
 Y1="Q.kplus1~L1a", L2a="Q.kplus1 ~ 1", L2b="Q.klus1~A1", Y2="Q.kplus1~A2 + gender"))
summary(result5c)
```

If there were a Anode or Cnode between L1b and Y1, Y1 would also need a Q regression formula.


